Maud: Bayesian regression modelling facilitates quantitative modelling of cell metabolism

### 2023-11-01 paper/main.tex
Implement Lars's paper comments

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Bayesian regression facilitates quantitative modelling of cell metabolism},
  pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
  pdfkeywords={Bayesian inference, kinetic models of cell metabolism,
multiomics integration, ordinary differential equations, regulatory
anlaysis},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Our paper reviews the existing options in this
area, presents a case study illustrating how Maud can be used to analyse
a metabolic network and explains the biological, statistical and
computational design decisions underpinning Maud.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

A kinetic model of cellular metabolism aims to express what is known
about a cellular process in the form of an \emph{in silico}
representation of the underlying network of chemical reactions. Kinetic
models can be used to improve production of target molecules, determine
regulatory networks \citep{christodoulou_reserve_2018} and identify
potential drug targets
\citep{deberardinis_fundamentals_2016, Liberti2017}. However, the use of
kinetic models in practice is hindered by their dependence on noisy and
uncertain information sources. Quantitative in vivo measurements of
chemical abundances, and in vitro measurements relating to kinetic
parameters, both contain vital information but are notoriously
inaccurate {[}siskos2017; tabb2010{]}. Practically useful kinetic
modelling therefore requires a principled statistical approach that
encompasses multiple possible model parameterisations.

@@

In equation \eqref{eq-1}, C represents a vector of metabolite
concentrations, \(v\) is a vector of reaction rates, and \(\mu\) is the
growth rate. The second term represents the dilution due to cell growth.

In a kinetic model, the rates, \(v\), are expressed as a function of the
enzyme concentrations, \(E\), the metabolite concentrations, \(C\), and
a set of parameters, \(\theta\) as shown in equation \eqref{eq-2}

@@

In a fermentation context, equation \eqref{eq-3} captures the rapid
kinetics inside the cell, while another set of ODEs would be used to
describe the external substrate and product concentrations, which could
act as boundary parameters to equation \eqref{eq-3}.

In the context of kinetic modelling, Bayesian inference is appealing
because it allows uncertainty to be represented appropriately without
sacrificing mechanistic accuracy. Measurement uncertainty can naturally
be represented in a Bayesian measurement model, whereas the prior model
can represent quantitative uncertainty about kinetic parameters.
Finally, kinetic rate laws can be represented in Bayesian data
generation models with arbitrarily high fidelity. See
\citet{gelmanBayesianDataAnalysis2020a} for more about Bayesian
inference and \citet{gelmanBayesianWorkflow2020} for a discussion of
practical Bayesian workflow.

Another advantage is that Bayesian inference problems are well-posed
even when not all parameters are strongly identified. Sloppy models in
which measurable quantities are sensitive to combinations of parameters
but not to individual marginal parameter values are ubiquitous in models
of biological systems \citep{gutenkunst_2007, white_2016}. The parameter
correlation structure represents the set of potential models that
describe the observed data. As we demonstrate in
Section~\ref{sec-laplace} below, capturing this correlation structure is
difficult outside a fully Bayesian context.

@@

The most popular algorithm for fitting Bayesian statistical models is
Markov Chain Monte Carlo (MCMC). Modern MCMC algorithms allow
exploration of high- dimensional posterior distributions, have robust
failure diagnostics
\citep{vehtariRankNormalizationFoldingLocalization2021} and can
incorporate fast numerical solvers, thereby making inference feasible
for Bayesian kinetic models. Nonetheless, the kinetic modelling
literature reports an aversion to MCMC, rooted mainly in concerns about
sampling time and the presumed difficulty of implementing the required
statistical model \citep{raue_joining_2013, saa_construction_2016}. We
are only aware of two recent attempts to implement a Bayesian kinetic
modelling approach using MCMC. \citet{stapor_pesto_2018} fitted detailed
kinetic models using relatively inefficient MCMC algorithms that do not
scale well to high dimensional parameter spaces limiting the scope of
modelling. Conversely, \citet{st.johnBayesianInferenceMetabolic2018}
utilises an efficient sampling algorithm but uses approximate kinetics,
namely lin-log kinetics \citet{visser_dynamic_2003}, limiting the scope
of interpreting parameters and inferring cellular behaviour in
experimental conditions outside the reference dataset.

There have also been efforts to implement Bayesian inference for kinetic
models without the use of MCMC. Examples of alternative inference
methods include variational inference
\citep{st.johnBayesianInferenceMetabolic2018}, rejection sampling and
approximate Bayesian computation \citep{saa_construction_2016} and
Laplace approximation, in which the Fisher information matrix is used to
calculate a normal approximation around the maximum a posteriori
parameter configuration
\citep{liebermeister_model_2021, gopalakrishnan_k-fit_2020, stapor_pesto_2018, raue_joining_2013}.
Non-MCMC-based Bayesian kinetic models have limited utility because they
lack reliable diagnostic tools for verifying that their results
approximate the target posterior distribution. This is a problem because
realistic kinetic models tend to induce highly correlated, non-Gaussian,
joint probability distributions
\citep{gutenkunst_2007, stapor_pesto_2018}.

@@

\includegraphics{../figures/methionine-reactions.png}

@@

We constructed a kinetic model of the methionine cycle in Maud's format
using the description in \citet{korendyaseva_allosteric_2008}. The
ordinary differential equation system describing this model is shown in
Equation\,\eqref{eq-meth-ode}.

@@

We used these parameters to simulate steady states in a range of
plausible experimental conditions, again using
\citet{saa_construction_2016} as a starting point. These steady states
were then used to generate simulated measurements using the measurement
model described below in Section~\ref{sec-methionine-measurement-model}.

\hypertarget{sec-methionine-measurement-model}{%
\subsubsection{Measurement
model}\label{sec-methionine-measurement-model}}

Generating an artificial dataset required a specification of the true
measurement error. For enzyme and metabolite concentration measurements
we specified a standard deviation of 0.1 on natural logarithmic scale,
corresponding to approximately 10\% measurement error. For each reaction
measurement a measurement standard deviation of approximately 10\% of
the simulated value.

@@

For our main model run, we assumed that all metabolite and enzyme
concentrations were measured, and that there was a reaction measurement
for each of the network's elementary flux modes.

\hypertarget{trainingvalidation-split}{%
\subsubsection{Training/validation
split}\label{trainingvalidation-split}}

The split between training and validation conditions was selected to
achieve a large difference in fluxes between the two categories. The
split was determined as we are interested in showing how our model can
fit to varied conditions, and conditions closer to the training set are
likely to be predicted well without necessarily learning the system.

\hypertarget{additional-dataset-with-missing-measurements}{%
\subsubsection{Additional dataset with missing
measurements}\label{additional-dataset-with-missing-measurements}}

\hypertarget{maud-input-specification}{%
\subsubsection{Maud input
specification}\label{maud-input-specification}}

@@

@@

\hypertarget{findings}{%
\subsection{Findings}\label{findings}}

\hypertarget{posterior-inference}{%
\subsubsection{Posterior inference}\label{posterior-inference}}

@@

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared with
the corresponding marginal prior distributions, even though these
parameters' joint posterior distribution contained enough information to
make accurate out of sample predictions. In some cases, there were
two-dimensional correlations such as the one shown in the bottom left of
Figure~\ref{fig-posterior}; in this case the marginal distribution of
the two parameters is roughly banana shaped. More commonly, however,
two-dimensional pair plots were insufficient to reveal the underlying
correlation structure, as seen in the bottom-right plot in figure
Figure~\ref{fig-posterior}. This does not mean that the parameters were
uncorrelated, but rather that the correlations involve more than two
parameters.

@@

The question naturally arises whether the crucial high-dimensional
parameter correlations are linear or non-linear. This is relevant to the
question of model performance, as linear correlations are easier to
correct for. A linearly correlated posterior space would also be easier
to summarise. We address this question below in
Section~\ref{sec-laplace}.

This case study illustrates the type of kinetic model and dataset that
Maud can fit. The model we analysed has 10 reactions, 5 state variables
and 212 parameters. Generalising from our ability to fit this model in a
reasonable time using Maud, we expect that Maud can be used to fit
realistic Bayesian models of approximately the same size, but not, for
example, genome-scale kinetic models. To fit larger models, faster
steady state solving methods or alternative inference algorithms will be
required. Section~\ref{sec-laplace} addresses whether Laplace
approximation is a suitable candidate.

@@

\includegraphics{../figures/posterior.png}

@@

\caption{\label{fig-posterior}Marginal posterior distributions from our
main model run. (\textbf{Top left}) Comparison of posterior predictive
intervals with simulated flux measurement values. All the flux
measurements are within the predictive intervals, indicating a good fit.
(\textbf{Top right}) Comparison of posterior predictive intervals with
simulated concentration measurement values. These also show a good fit.
(\textbf{Bottom left}) Pairwise marginal posterior distribution for two
uncorrelated parameters, namely \(K_m^{AHC1,hcys-L}\) and
\(K_m^{AHC1_ahcys}\). (\textbf{Bottom right}) Pairwise marginal
posterior distribution for two correlated parameters, namely
\(K_m^{MAT3,atp}\) and \(K_{cat}^{MAT3}\).}

@@

We found that the Laplace method was not able to produce an accurate
posterior approximation for our model and dataset.

Our overall strategy was to compare samples generated using MCMC, which
we were confident could be treated as draws from the true posterior
distribution, with approximate posterior samples generated using the
Laplace method. We were unable to generate approximate posterior samples
for our main methionine cycle case study using Laplace approximation, as
the algorithm could not recover from solver failures caused by
unrealistic parameter configurations. We therefore made a comparison for
a simpler model. This input can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/data/example_ode}.

MCMC sampling for our comparison input yielded 800 samples in 625
minutes; Laplace sampling yielded the same number of samples in only one
minute. The diagnostics indicated that our algorithm was able to find
the maximum a posteriori parameter configuration, approximate the
Hessian and use these quantities to generate approximate posterior
samples. The results can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/results}.

Figure~\ref{fig-laplace} summarises the results of comparing the samples
generated using each method. As can be seen from the top left plot, the
Laplace method does not provide a good approximation to the true
posterior distribution in this case, as the marginal distribution of the
total log probability density is clearly different. This impression was
confirmed using the Kolmogorov-Smirnov test, which is a test to
differentiate two empirical univariate distributions. The two
distributions were significantly different with a p-value
indistinguishable from zero.

The difference between the Laplace approximation output and the true
posterior distribution manifests not only in the parameter space, but
also in the measurement space. The upper right (training data) and lower
left (validation data) plots in Figure~\ref{fig-laplace} show 5\%-95\%
marginal flux measurement log likelihood intervals for both methods;
lower log likelihood values indicate that the modelled and measured
values are further away. The graph shows that the Laplace approximation
yielded significantly worse predictions than the true posterior, even
for the training data.

To further explore why this is the case we compared samples from the
true posterior and the Laplace approximation for the pairwise marginal
distributions of two Michaelis-Menten constants \(K_{m}^{A,r1}\) and
\(K_{cat}^{r1}\): see the bottom right cell of Figure~\ref{fig-laplace}.
This comparison demonstrates that the Laplace method is not able to
capture the correct relationships between parameters' distributions.

This result shows that MCMC, while slower than Laplace approximation, is
unfortunately preferable for posterior inference in this case. We expect
that the Laplace method will produce worse approximations the more
complex the target model. Since the approximation is already
unacceptable for our simple test model, we recommend that Maud users use
MCMC sampling in preference to Laplace approximation if possible when
fitting realistic Bayesian kinetic models.

@@

\includegraphics{../figures/laplace.png}

@@

\caption{\label{fig-laplace}Graphical comparison of approximate
posterior samples generated using Laplace sampling (blue-green) with
posterior samples generated using MCMC (dark grey). (\textbf{Top left})
the two sets of samples clearly have different marginal distributions
for the overall log probability variable, indicating that the Laplace
samples do not accurately approximate the target distribution.
(\textbf{Top right}) The distribution of marginal posterior predictive
log likelihood values for training data flux measurements shows that the
Laplace method tended to yield much worse predictions compared with the
true model. (\textbf{Bottom left}) The Laplace method also tended to
produce worse flux predictions for held-back test measurements.
(\textbf{Bottom right}) The marginal joint distribution of two
parameters: \(K_m^{A,rl}\) and \(K_{cat}^{r1}\). The Laplace method is
not able to track the correct joint distribution for this pair of
parameters. This is unsurprising given that the target distribution has
position-dependent scales which are difficult for a linear approximation
to capture.}

@@

\hypertarget{effect-of-missing-metabolite-concentration-measurements}{%
\subsubsection{Effect of missing metabolite concentration
measurements}\label{effect-of-missing-metabolite-concentration-measurements}}

@@

Figure~\ref{fig-missing} shows that, as might be expected, the model
with missing measurements did not correctly infer the missing ahcys
concentrations. Nonetheless, the remaining measured metabolites were
still well predicted, suggesting that information about the network is
still preserved despite the missing measurements. Comparison of flux
measurements in both models also indicated that removing the ahcys
measurement did not result in catastrophic model failure.

The missing measurements did affect Maud's ability to infer parameter
values correctly. The lower left plot of Figure~\ref{fig-missing} shows
that the model with full metabolomics learned the true value for the
displayed dissociation constant, despite this value being far from the
mean of the corresponding marginal prior distribution. In contrast, the
model with missing measurements stayed in the neighbourhood of the
prior.

@@

\includegraphics{../figures/missing.png}

@@

\hypertarget{application-to-regulatory-understanding}{%
\subsubsection{Application to regulatory
understanding}\label{application-to-regulatory-understanding}}

To demonstrate how Maud's output can be used to yield useful metabolic
insights we used the results of our case study to explain why the flux
of the enzyme GNMT is higher in dataset 1 than in dataset 12. GNMT is an
irreversible enzyme that is homotropically activated by its substrate,
competitively inhibited by its product and heterotropically inhibited by
5,10- methylenetetrahydrofolate (mlthf). The diverse regulation makes it
the ideal test case to elucidate regulatory changes.

Figure~\ref{fig-decomposition} shows the regulatory description of GNMT,
according to the results our main case study analysis. Each curve shows
the marginal posterior distribution of the ratio of the corresponding
regulatory component in dataset 1 compared with dataset 12. A positive
value indicates that the component was increased in dataset 1 relative
to dataset 12, with 0 indicating no difference. The probability,
according to our model, of the component acting in each direction is
given by the relative area under the curve on each side of the zero
point.

@@

\includegraphics{../figures/decomposition.png}

@@

\caption{\label{fig-decomposition}Illustration of how analysing a system
with Maud can yield actionable insights about the underlying metabolic
network. (\textbf{Top Left}) A schematic of the regulatory interactions
associated with the enzyme \(GNMT1\). Dashed green lines represents
allosteric activation, dashed red lines indicate allosteric inhibition
and solid red lines represent competitive inhibition. (\textbf{Bottom
Left}) Comparison of marginal posterior distributions for \(GNMT\) flux
in datasets 1 and 2. (\textbf{Right}) Log-scale ratios of the regulatory
elements defined in equation \eqref{eq-decomposition}. Note that the
\(Reversibility\) and \(k_cat\) components are excluded: this is because
this reaction was modelled as irreversible, and \(k_{cat}\) was modelled
as constant across datasets. These plots identify why flux in dataset 12
is higher than in dataset 1: the flux increase is due to allostery and
saturation with no control from enzyme concentration changes.}

@@

Maud's kinetic model decomposes into factors as shown in equation
\eqref{eq-decomposition}.

>>>>

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Maud: Bayesian regression modelling facilitates quantitative modelling of cell metabolism},
  pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
  pdfkeywords={Bayesian inference, kinetic models of cell metabolism,
multiomics integration, ordinary differential equations, regulatory
anlaysis},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\title[]{Maud: Bayesian regression modelling facilitates quantitative
modelling of cell metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Our paper reviews the existing options in this
area, presents a case study illustrating how Maud can be used to analyse
a metabolic network and explains the biological, statistical and
computational design decisions underpinning Maud.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, interior hidden, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

A kinetic model of cellular metabolism aims to express what is known
about a cellular process in the form of an \emph{in silico}
representation of the underlying network of chemical reactions. Kinetic
models can be used to improve production of target molecules, determine
regulatory networks \citep{christodoulou_reserve_2018} and identify
potential drug targets
\citep{deberardinis_fundamentals_2016, Liberti2017}. However, the use of
kinetic models in practice is hindered by their dependence on noisy and
uncertain information sources. Quantitative in vivo measurements of
chemical abundances, and in vitro measurements relating to kinetic
parameters, both contain vital information but are notoriously
inaccurate \citep{siskos2017, tabb2010, lu2017}. Practically useful
kinetic modelling therefore requires a principled statistical approach
that encompasses multiple possible model parameterisations.

@@

where C represents a vector of metabolite concentrations, \(v\) is a
vector of reaction rates, and \(\mu\) is the growth rate. The second
term represents the dilution due to cell growth.

In a kinetic model, the rates, \(v\), are expressed as a function of the
enzyme concentrations, \(E\), the metabolite concentrations, \(C\), and
a set of parameters \(\theta\):

@@

In a fermentation context, \eqref{eq-3} captures the rapid kinetics
inside the cell, while another set of ODEs would be used to describe the
external substrate and product concentrations, which could act as
boundary parameters to \eqref{eq-3}.

In the context of kinetic modelling, Bayesian inference is appealing
because it allows uncertainty to be represented appropriately without
sacrificing mechanistic accuracy. Measurement uncertainty can naturally
be represented in a Bayesian measurement model, whereas the prior model
can represent quantitative uncertainty about kinetic parameters.
Finally, kinetic rate laws can be represented in Bayesian data
generation models with arbitrarily high fidelity. See
\citet{gelmanBayesianDataAnalysis2020a} for more details about Bayesian
inference and \citet{gelmanBayesianWorkflow2020} for a discussion of
practical Bayesian workflow.

Another advantage is that Bayesian inference problems are well-posed
even when not all parameters are strongly identified. Sloppy parameter
models in which measurable quantities are sensitive to combinations of
parameters but not to individual marginal parameter values are
ubiquitous in models of biological systems
\citep{gutenkunst_2007, white_2016}. The parameter correlation structure
represents the set of potential models that describe the observed data.
As we demonstrate later, capturing this correlation structure is
difficult outside a fully Bayesian context.

@@

The most popular algorithm for fitting Bayesian statistical models is
Markov Chain Monte Carlo (MCMC). Modern MCMC algorithms enable
exploration of high-dimensional posterior distributions, have robust
failure diagnostics
\citep{vehtariRankNormalizationFoldingLocalization2021} and can
incorporate fast numerical solvers, thereby making inference feasible
for Bayesian kinetic models. Nonetheless, the kinetic modelling
literature reports an aversion to MCMC, rooted mainly in concerns about
sampling time and the presumed difficulty of implementing the required
statistical model \citep{raue_joining_2013, saa_construction_2016}. We
are only aware of two recent attempts to implement a Bayesian kinetic
modelling approach using MCMC. \citet{stapor_pesto_2018} fitted detailed
kinetic models using relatively inefficient MCMC algorithms that do not
scale well to high dimensional parameter spaces limiting the scope of
modelling. Conversely, \citet{st.johnBayesianInferenceMetabolic2018}
utilises an efficient sampling algorithm but uses approximate kinetics,
namely lin-log kinetics \citep{visser_dynamic_2003}, limiting the scope
of interpreting parameters and inferring cellular behaviour in
experimental conditions outside the reference dataset.

There have also been efforts to implement Bayesian inference for kinetic
models without the use of MCMC. Examples of alternative inference
methods include variational inference
\citep{st.johnBayesianInferenceMetabolic2018}, rejection sampling and
approximate Bayesian computation \citep{saa_construction_2016} and
Laplace approximation, in which the Fisher information matrix is used to
calculate a normal approximation around the maximum a posteriori
parameter configuration
\citep{liebermeister_model_2021, gopalakrishnan_k-fit_2020, stapor_pesto_2018, raue_joining_2013}.
Non-MCMC-based Bayesian kinetic models suffer from a lack of reliable
diagnostic tools for verifying that their results approximate the target
posterior distribution. This is a problem because realistic kinetic
models tend to induce highly correlated, non-Gaussian, joint probability
distributions \citep{gutenkunst_2007, stapor_pesto_2018}.

@@

\includegraphics{./figures/methionine-reactions.png}

@@

We constructed a kinetic model of the methionine cycle in Maud's format
using the description in \citet{korendyaseva_allosteric_2008}. The
ordinary differential equation system describing this model is

@@

We used these parameters to simulate steady states in a range of
plausible experimental conditions, again using
\citet{saa_construction_2016} as a starting point. These steady states
were then used to generate simulated measurements using the measurement
model.

For enzyme and metabolite concentration measurements we specified a
standard deviation of 0.1 on natural logarithmic scale, corresponding to
approximately 10\% measurement error. For each reaction measurement a
measurement standard deviation of approximately 10\% of the simulated
value.

@@

For our main model run, we assumed that all metabolite and enzyme
concentrations were measured, and that there was a reaction measurement
for each of the network's free fluxes.

The simulated experiments and measurements were split into a training
and a validation dataset in a way that achieved a large difference in
flux between the two categories. This was done to evaluate whether the
fitted model is able to extrapolate to conditions well outside the
training dataset rather than merely interpolating between the training
data without necessarily learning the system.

@@

\hypertarget{posterior-inference}{%
\subsection{Posterior inference}\label{posterior-inference}}

@@

@@

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared to the
corresponding marginal prior distributions, even though these
parameters' joint posterior distribution contained enough information to
make accurate out of sample predictions. In some cases, there were
two-dimensional correlations such as the one shown in the bottom left of
Figure~\ref{fig-posterior}; in this case the marginal distribution of
the two parameters is roughly banana shaped. More commonly, however,
two-dimensional pair plots were insufficient to reveal the underlying
correlation structure, as seen in the bottom-right plot in figure
Figure~\ref{fig-posterior}, which shows two marginally independent
parameters. This kind of marginal independence does not mean that the
parameters are independent in the full joint posterior distribution.

@@

The question naturally arises whether the crucial high-dimensional
parameter correlations are linear or non-linear. This is relevant to the
question of model performance, as linear correlations are easier to
correct for. A linearly correlated posterior space would also be easier
to summarise. We address this question in the next section.

@@

\includegraphics{./figures/posterior.png}

@@

\caption{\label{fig-posterior}Marginal posterior distributions from our
main model run. (\textbf{Top left}) Comparison of posterior predictive
intervals with simulated flux measurement values. All the flux
measurements are within the predictive intervals, indicating a good fit.
(\textbf{Top right}) Comparison of posterior predictive intervals with
simulated concentration measurement values. These also show a good fit.
(\textbf{Bottom left}) Pairwise marginal posterior distribution for two
pairwise correlated parameters, namely \(K_m^{AHC1,hcys-L}\) and
\(K_m^{AHC1_ahcys}\). (\textbf{Bottom right}) Pairwise marginal
posterior distribution for two pairwise-uncorrelated parameters, namely
\(K_m^{MAT3,atp}\) and \(K_{cat}^{MAT3}\).}

@@

This current case study illustrates the type of kinetic model and
dataset that Maud can fit. The model we analysed has 10 reactions, 5
state variables and 212 parameters. Generalising from our ability to fit
this model in a reasonable time using Maud, we expect that Maud can be
used to fit realistic Bayesian models of approximately the same size,
but not, for example, genome-scale kinetic models. To fit larger models,
faster steady state solving methods or alternative inference algorithms
will be required. The Laplace approximation, in which the Fisher
information matrix is used to calculate a normal approximation around
the maximum a posteriori parameter configuration, is a popular strategy
for generating approximate posterior samples while avoiding a full MCMC
approach.

Using the recently implemented Laplace approximation in Stan, we were
unable to generate approximate posterior samples for our main methionine
cycle case study using Laplace approximation, as the algorithm could not
recover from solver failures caused by unrealistic parameter
configurations. This is not an inherent issue of the Laplace
transformation but highlights some of the challenges around solving
\eqref{eq-3} for real problems.

In order still to assess the potential use of the Laplace approximation,
we made a comparison for a simpler model. This input can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/data/example_ode}.
MCMC sampling for this simpler model yielded 800 samples in 625 minutes,
while Laplace sampling yielded the same number of samples in only one
minute. The diagnostics indicated that our algorithm was able to find
the maximum a posteriori parameter configuration, approximate the
Hessian and use these quantities to generate approximate posterior
samples. The results can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/results}.

Comparing the samples generated using each method shows that the Laplace
approximation does not provide a good approximation to the true
posterior distribution in this case (Figure~\ref{fig-laplace}). As can
be seen from the top left plot, the total log probability density is
clearly different and this was confirmed by a test of the equality of
two empirical univariate distributions (Kolmogorov-Smirnov,
\(p=1.7\times10^{-65}\)).

The difference between the Laplace approximation output and the true
posterior distribution manifests not only in the parameter space, but
also in the measurement space for both training data (upper right) and
validation data (lower left). The lower log likelihood values indication
that the modelled and measured values are further away and that the
Laplace approximation yielded significantly worse predictions than the
true posterior, even for the training data.

To further explore why this is the case we compared samples from the
true posterior and the Laplace approximation for the pairwise marginal
distributions of two Michaelis-Menten constants \(K_{m}^{A,r1}\) and
\(K_{cat}^{r1}\) (Figure~\ref{fig-laplace}, bottom right). This
comparison demonstrates that the Laplace method is not able to capture
the correct relationships between parameters' distributions.

This result shows that MCMC, while slower than Laplace approximation, is
required for posterior inference in this case. We expect that the
Laplace method will produce worse approximations the more complex the
target model. Since the approximation is already unacceptable for our
simple test model, we recommend that Maud users use MCMC sampling in
preference to Laplace approximation if possible when fitting realistic
Bayesian kinetic models.

@@

\includegraphics{./figures/laplace.png}

@@

\caption{\label{fig-laplace}Graphical comparison of approximate
posterior samples generated using Laplace sampling (blue-green) with
posterior samples generated using MCMC (dark grey). (\textbf{Top left})
the two sets of samples clearly have different marginal distributions
for the overall log probability variable, indicating that the Laplace
samples do not accurately approximate the target distribution.
(\textbf{Top right}) The distribution of marginal posterior predictive
log likelihood values for training data flux measurements shows that the
Laplace method tended to yield much worse predictions compared with the
true model (lower log likelihood values indicate that the modelled and
measured values are further away). (\textbf{Bottom left}) The Laplace
method also tended to produce worse flux predictions for held-back test
measurements. (\textbf{Bottom right}) The marginal joint distribution of
two parameters: \(K_m^{A,rl}\) and \(K_{cat}^{r1}\). The Laplace method
is not able to track the correct joint distribution for this pair of
parameters. This is unsurprising given that the target distribution has
position-dependent scales which are difficult for a linear approximation
to capture.}

@@

\hypertarget{effect-of-missing-metabolite-concentration-measurements}{%
\subsection{Effect of missing metabolite concentration
measurements}\label{effect-of-missing-metabolite-concentration-measurements}}

@@

As might be expected, the model with missing measurements did not
correctly infer the missing ahcys concentrations
(Figure~\ref{fig-missing}). Nonetheless, the remaining measured
metabolites were still well predicted, suggesting that information about
the network is still preserved despite the missing measurements.
Comparison of flux measurements in both models also indicated that
removing the ahcys measurement did not result in catastrophic model
failure.

The missing measurements did affect Maud's ability to infer parameter
values correctly (Figure~\ref{fig-missing}, lower left). The model with
full metabolomics learned the true value for the displayed dissociation
constant, despite this value being far from the mean of the
corresponding marginal prior distribution. In contrast, the model with
missing measurements stayed in the neighbourhood of the prior.

@@

\includegraphics{./figures/missing.png}

@@

\hypertarget{application-to-regulatory-understanding}{%
\subsection{Application to regulatory
understanding}\label{application-to-regulatory-understanding}}

To demonstrate how Maud's output can be used to yield useful metabolic
insights we used the results of our case study to explain why the flux
of the enzyme GNMT is higher in dataset 1 than in dataset 12
(Figure~\ref{fig-decomposition}). GNMT is an irreversible enzyme that is
homotropically activated by its substrate, competitively inhibited by
its product and heterotropically inhibited by 5,10-
methylenetetrahydrofolate (mlthf). The complex regulation makes it the
ideal test case to elucidate regulatory changes.

Regulation can be separated into enzyme abundance, allostery and
saturation, and we can plot the marginal posterior distribution of the
ratio of the corresponding regulatory component in dataset 1 compared
with dataset 12 (Figure~\ref{fig-decomposition}, right panel). A
positive value indicates that the component was increased in dataset 1
relative to dataset 12, with 0 indicating no difference. The
probability, according to our model, of the component acting in each
direction is given by the relative area under the curve on each side of
the zero point.

@@

\includegraphics{./figures/decomposition.png}

@@

\caption{\label{fig-decomposition}Illustration of how analysing a system
with Maud can yield actionable insights about the underlying metabolic
network. (\textbf{Top Left}) A schematic of the regulatory interactions
associated with the enzyme \(GNMT1\). Dashed green lines represents
allosteric activation, dashed red lines indicate allosteric inhibition
and solid red lines represent competitive inhibition. (\textbf{Bottom
Left}) Comparison of marginal posterior distributions for \(GNMT\) flux
in datasets 1 and 2. (\textbf{Right}) Log-scale ratios of the regulatory
elements defined in \eqref{eq-decomposition}. Note that the
\(Reversibility\) and \(k_cat\) components are excluded: this is because
this reaction was modelled as irreversible, and \(k_{cat}\) was modelled
as constant across datasets. These plots identify why flux in dataset 12
is higher than in dataset 1: the flux increase is due to allostery and
saturation with no control from enzyme concentration changes.}

@@

Maud's kinetic model decomposes into factors:
